Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-12 writeup/conStruct.tex
g's changes

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed.
Although \texttt{conStruct}'s computation time is invariant 
with the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it scales poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which is an operation of order $O(n^3)$, 
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples. \gb{is this true?}
The other main difference is the model of noise,
but it is unclear what effects this may have,
or indeed, which is closer to the truth.

\paragraph{Choosing the ``best" number of clusters}
Although we recognize the utility of choosing a single, ``best'' value of $K$, 
and using only that analysis to communicate results, 
we want to emphasize that there are both statistical and biological reasons 
why there is no true best value of $K$.
From a statistical perspective, the choice of best $K$ is always relative to the data in hand.
This means both that sampling different individuals or loci 
from the same populations may yield a different ``best" $K$, 
but also 
-- unless the data were generated under the model itself --
that the support for larger values of $K$ is likely to increase with increasing amounts of data
(although perhaps only logarithmically).
In the limit of infinite data, the best value of $K$ 
may be the number of individuals included in the dataset.
From a biological perspective, 
it is important to stress that patterns of relatedness between individuals and populations 
are shaped by complex spatial and hierarchical processes.
All individuals within a species (and indeed, all individuals across all species), 
are related to one another in some way, 
and summarizing that relatedness with a single value of $K$ may be reductive or misleading.
We therefore encourage users to perform analyses across different values of $K$ and 
observe which layers split out at what levels 
(effectively taking successively shallower cross-sections of the population phylogeny),
and also to take the results of the proposed cross-validation procedure with a large grain of salt.
Calculating layer importance may also be a useful heuristic, 
as it can reveal layers with strong statistical support but small biological import.

\paragraph{Implications for management and conservation}
If isolation by distance is common, 
then a likely result of applying \texttt{conStruct} to existing data is that 
populations previously identified using nonspatial clustering methods 
will be collapsed into each other.  
This ``lumping" might better reflect biological reality, 
but may also have implications for management decisions and conservation policy, 
both of which are often predicated on the identification of discrete ``management units" (MUs) 
identified using genetic data \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.

It is therefore important to emphasize that individuals sampled from the same \texttt{conStruct} layer 
may be quite genetically diverged from one another, 
and that a \texttt{conStruct} layer may still contain multiple distinct MUs.  
Alternatively, the collapse of multiple, previously identified MUs into a single \texttt{conStruct} layer 
may reflect a biological reality that these populations are currently 
(or were recently) exchanging migrants, 
and thus might emphasize the importance of maintaining habitat corridors between demes, 
or of implementing an integrated conservation plan across multiple demes within a layer.

\paragraph{Caveats and considerations}
There are a few important caveats to consider in the interpretation of \texttt{conStruct} results. 
First, we have modeled allelic covariance within a layer as a spatial process.
Although there is flexibility built into the model about the shape of that covariance, 
inference may be misleading if the sampling geography departs radically from the way 
the sampled organisms experience (or have experienced) their landscape.
For example, if we were to run a \texttt{conStruct} analysis using geographic distances between 
sampled individuals of greenish warblers \citep{Irwin2001} or \textit{Ensatina} salamanders \citep{wake_schneider1998} 
--- two canonical examples of rings species --- 
we might get misleading results.
This is because distance between locations on either side of the species' distributions
(across the Tibetan plateau and the Central Valley, respectively) 
is not representative of the path traversed in the coalescent of a pair of alleles sampled at those locations.

@@

>>>>

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed.
Although \texttt{conStruct}'s computation time is invariant 
with the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it \gc{currently scales} poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which is an operation of order $O(n^3)$, 
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples. \gb{is this true?}
\gc{Our speed, on large sample sizes, could be improved by adopting rank one updates to the
 inverse of the covariance matrix for updating an individual's
 admixture proportions, which alters only a single row/column of the
 covariance matrix. However, we have not implemented this yet as it
 would likely mean losing our Hamilton MCMC for our parameters.}

%The other main difference is the model of noise,
%but it is unclear what effects this may have,

\gc{SO I THINK IF YOU SAY THIS you need to give some run time specs,
  mainly to show that it's not insanely slow.}

%Binomially clearly is, genotypes are discrete:  "or indeed, which is closer to the truth."

\paragraph{Choosing the ``best" number of clusters}
Although we recognize the utility of choosing a single, ``best'' value of $K$, 
and using only that analysis to communicate results, 
\gc{we emphasize that there} is no true best value of $K$.
\gc{The choice} of best $K$ is always relative to the data in hand.
%This means both that sampling different individuals or loci 
%from the same populations may yield a different ``best" $K$, 
From the statistical perspective, Unless the data were generated under the model itself 
that the support for larger values of $K$ is likely to increase with increasing amounts of data
%(although perhaps only logarithmically).
In the limit of infinite data, the best value of $K$ 
may be the number of samples included in the dataset \citep{Patterson2006}.
From a biological perspective, 
it is important to stress that patterns of relatedness between individuals and populations 
are shaped by complex spatial and hierarchical processes.
All individuals within a species (and indeed, all individuals across all species), 
are related to one another in some way, 
and summarizing that relatedness with a single value of $K$ may be reductive or misleading.
We therefore encourage users to perform analyses across different values of $K$ and 
observe which layers split out at what levels 
(\gc{conceptually similar to} taking successively shallower
cross-sections of the population phylogeny),  %%note that the
                                %%equivalent would only be ~true if we
                                %%had same sample size from each
                                %%proportion.
and also to take the results of the proposed cross-validation procedure with a large grain of salt.
Calculating layer importance may also be a useful heuristic, 
as it can reveal layers with statistical support but small biological import.

\gc{In adding to spatial realism to the groups used by clustering methods,
it is important to re-emphasize that that clusters/layers detected by
these methods do not necessarily accord to distinct, ancestral
populations. Nor does a non-zero admixture proportion indicate that admixture (i.e. gene
flow) must have occurred. Both groupings and admixture proportions
should be viewed as hypotheses that should be subject to futher
testing, see \citet{Falush:16} for futher discussion of these points. }

\paragraph{Implications for management and conservation}
If isolation by distance is common, 
then a likely result of applying \texttt{conStruct} to existing data is that 
populations previously identified using nonspatial clustering methods 
will be collapsed into each other.  
This ``lumping" might better reflect biological \gc{``reality'',  
but may also have implications for management decisions and conservation policy, 
both of which are often predicated on the identification of discrete ``management units" (MUs) 
identified using genetic data \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.

It is therefore important to emphasize that individuals sampled from the same \texttt{conStruct} layer 
may be quite genetically diverged from one another, \gc{perhaps especially at
  loci underlying adaptive traits}, 
and that a \texttt{conStruct} layer may still contain multiple distinct MUs.  
Alternatively, the collapse of multiple, previously identified MUs into a single \texttt{conStruct} layer 
may reflect a biological reality that these populations are currently 
(or were recently) exchanging migrants, 
and thus might emphasize the importance of maintaining habitat corridors between demes, 
or of implementing an integrated conservation plan across multiple demes within a layer.

\paragraph{Caveats and considerations}
There are a few important caveats to consider in the interpretation of \texttt{conStruct} results. 
First, we have modeled allelic covariance within a layer as a spatial process.
Although there is flexibility built into the model about the shape of that covariance, 
inference may be misleading if the sampling geography departs radically from the way 
the sampled organisms experience (or have experienced) \gc{dispersal
  on } their landscape.
For example, if we were to run a \texttt{conStruct} analysis using geographic distances between 
sampled individuals of greenish warblers \citep{Irwin2001} or \textit{Ensatina} salamanders \citep{wake_schneider1998} 
--- two canonical examples of rings species --- 
we might get misleading results.
This is because distance between locations on either side of the species' distributions
(across the Tibetan plateau and the Central Valley, respectively) 
is not representative of the path traversed in the coalescent of a pair of alleles sampled at those locations.

@@

%\gc{We could imagine modeling }
